A Fundamental Measure Theory Functional for Hard-Sphere Contact Densities

### 2011-06-26 contactpaper/paper.tex
implement plot of contact density vs filling fraction and free energy
Tested-on: collins (Linux 2.6.32-5-amd64)
derive and discuss homogeneous limit of White Bear free energy

@@

\begin{align}
  \frac{dn_3(\mathbf{r})}{dR} &= -n_2(\mathbf{r})\\
  \frac{dn_2(\mathbf{r})}{dR} &= \int n(\mathbf{r}+\mathbf{r}')
  \delta'(|\mathbf{r'}| - R)\mathbf{dr}'\\
  &= \int n(\mathbf{r}+\mathbf{r}')
  \delta'(r' - R) d\Omega' r'^2 dr'\\
  &= \int 
  \delta(r' - R) d\Omega' \left(\frac{dn(\mathbf{r}+\mathbf{r}')}{dr'}r'^2 + 2n(\mathbf{r}+\mathbf{r}')r' \right) dr'\\
  &= \frac1{R} \int 
  \delta(|\mathbf{r}'| - R)
  \left(\mathbf{r}'\cdot \nabla n(\mathbf{r}+\mathbf{r}') +
  2n(\mathbf{r}+\mathbf{r}') \right) \mathbf{dr}' \\
  &= \frac{2}{R}n_2(\mathbf{r}) + \frac{1}{R} \int 
  \delta(|\mathbf{r}'| - R)\mathbf{r}'\cdot \nabla n(\mathbf{r}+\mathbf{r}') \mathbf{dr}' \\
  \frac{dn_0(\mathbf{r})}{dR} &= \frac{1}{R} \int 
  \delta(|\mathbf{r}'| - R)\mathbf{r}'\cdot \nabla n(\mathbf{r}+\mathbf{r}') \mathbf{dr}' \\
  \frac{d\mathbf{n}_{V2}}{dR} &= \nabla n_2
\end{align}

The total force due to $\Phi_1$ is actually very easy to work out:
\begin{align}
  \frac{d A_{HS}^{(1)}}{d R} &=
  \int \mathbf{dr}' \left(-\frac{n_0n_2}{1-n_3} - \frac{dn_0(\mathbf{r})}{dR}\ln(1-n_3) \right)
\end{align}
where you can see that $\frac{dn_0(\mathbf{r})}{dR}$ is zero for a
homogeneous system, and small for reasonably smooth density
distributions.  The next term from $\Phi_2$ is as follows:
\begin{align}
  \frac{d A_{HS}^{(2)}}{d R} &=
  \int \mathbf{dr}' \left(
    -\frac{n_1 n_2 - \mathbf{n}_{V1} \cdot\mathbf{n}_{V2}}{(1-n_3)^2}
    n_2
    - \frac{n_0n_2 - \mathbf{n}_{V0} \cdot\mathbf{n}_{V2} - 2 n_1
      \frac{dn_2}{dR} + 2 \mathbf{n}_{V1} \cdot \nabla n_2 \cdot
    }{1-n_3}
  \right) \\
  &= \int \mathbf{dr}' \left(
    -\frac{n_1 n_2 - \mathbf{n}_{V1} \cdot\mathbf{n}_{V2}}{(1-n_3)^2}
    n_2
    - \frac{n_0n_2 - \mathbf{n}_{V0} \cdot\mathbf{n}_{V2}
      - 2 n_1 \left(\frac2{R}n_2 \frac{dn_0}{dR}\right)
      + 2 \mathbf{n}_{V1} \cdot \nabla n_2 \cdot }{1-n_3}
  \right) \\
  &= \int \mathbf{dr}' \left(
    -\frac{n_1 n_2 - \mathbf{n}_{V1} \cdot\mathbf{n}_{V2}}{(1-n_3)^2}
    n_2
    - \frac{n_0n_2 - \mathbf{n}_{V0} \cdot\mathbf{n}_{V2}
      - 4 n_0n_2 - 2 n_1\frac{dn_0}{dR}
      + 2 \mathbf{n}_{V1} \cdot \nabla n_2 \cdot }{1-n_3}
  \right) \\
  &= \int \mathbf{dr}' \left(
    -\frac{n_1 n_2 - \mathbf{n}_{V1} \cdot\mathbf{n}_{V2}}{(1-n_3)^2}
    n_2
    + \frac{3n_0n_2 + \mathbf{n}_{V0} \cdot\mathbf{n}_{V2}
      + 2 n_1\frac{dn_0}{dR}
      - 2 \mathbf{n}_{V1} \cdot \nabla n_2 \cdot }{1-n_3}
  \right)
\end{align}

And the last term looks like this.  I should perhaps note that this
term should be replaced by a tensor version if want to examine
seriously localized density distributions.
\begin{align}
  A_{HS}^{(3)} &=
  \int \mathbf{dr}'\left(
     (n_2^3 - 3n_2\mathbf{n}_{V2} \cdot \mathbf{n}_{V2})
     \frac{
       n_3 + (1-n_3)^2 \ln(1-n_3)
     }{
       36\pi n_3^2\left( 1 - n_3 \right)^2
     }
  \right) \\
  \frac{d A_{HS}^{(3)}}{d R} &=
  \int \mathbf{dr}'
     \left[
       3(n_2^2 - \mathbf{n}_{V2} \cdot \mathbf{n}_{V2})
       \left(\frac2{R}n_2 + \frac{dn_0}{dR}\right)
       - 6n_2 \mathbf{n}_{V2} \cdot \nabla n_2
     \right]
     \frac{
       n_3 + (1-n_3)^2 \ln(1-n_3)
     }{
       36\pi n_3^2\left( 1 - n_3 \right)^2
     }
   \\
  &+
     (n_2^3 - 3n_2\mathbf{n}_{V2} \cdot \mathbf{n}_{V2}) (-n_2)
     \frac{
       (1 - 2(1-n_3)\ln(1-n_3) - (1-n_3))n_3^2( 1 - n_3)^2
       -
       (n_3 + (1-n_3)^2 \ln(1-n_3))(-2 n_3^2( 1 - n_3) + 2n_3(1-n_3)^2)
     }{
       36\pi (n_3^2\left( 1 - n_3 \right)^2)^2
     }
     \\
     &=
     \int \mathbf{dr}'
     \left[
       (n_1n_2 - \mathbf{n}_{V1} \cdot \mathbf{n}_{V2})
       \left(2n_2 + \frac{1}{4\pi}\frac{dn_0}{dR}\right)
       - \frac{1}{2\pi}n_2 \mathbf{n}_{V2} \cdot \nabla n_2
     \right]
     \frac{
       n_3 + (1-n_3)^2 \ln(1-n_3)
     }{
       3 n_3^2\left( 1 - n_3 \right)^2
     }
   \\
  &-
     \frac{n_2(n_2^3 - 3n_2\mathbf{n}_{V2} \cdot \mathbf{n}_{V2})}{36\pi}
   \left( \frac{1}{(1-n_3)^2} + 2 \frac{\ln(1-n_3)}{n_3^2(1-n_3)} \right)
\end{align}

@@

\begin{align}
    \beta\frac{\delta A_{HS}}{\delta n_3(\mathbf{r}')} &=
    -\frac{n_0(\mathbf{r}')}{1 - n_3(\mathbf{r}')}
    + \frac{n_1n_2 - \mathbf{n}_{V1}\cdot\mathbf{n}_{V2}}{(1 -
      n_3(\mathbf{r}'))^2}
    + \frac{n_2^3 -
      3n_2\mathbf{n}_{V2}\cdot\mathbf{n}_{V2}}{36\pi}\left(
      \frac{1}{(1-n_3)^2} + 2\frac{\ln(1-n_3)}{n_2^2(1-n_3)}
    \right)
\end{align}
\end{widetext}

@@

>>>>

\begin{figure}
\includegraphics[width=5cm]{figs/gHS-vs-n}
\caption{A comparison of the various approximations to the contact
  density.}
\label{fig:gHS-vs-n}
\end{figure}

\begin{figure}
\includegraphics[width=5cm]{figs/free-energy}
\caption{A comparison of the various approximations to the free
  energy.  Note that the White Bear functional uses a modified form of
  the Carnahan equation of state.}
\label{fig:free-energy}
\end{figure}

@@

\begin{align}
  \frac{dn_3(\mathbf{r})}{dR} &= n_2(\mathbf{r})\\
  \frac{dn_2(\mathbf{r})}{dR} &= \int n(\mathbf{r}+\mathbf{r}')
  \delta'(|\mathbf{r'}| - R)\mathbf{dr}'\\
  &= \int n(\mathbf{r}+\mathbf{r}')
  \delta'(r' - R) d\Omega' r'^2 dr'\\
  &= \int 
  \delta(r' - R) d\Omega' \left(\frac{dn(\mathbf{r}+\mathbf{r}')}{dr'}r'^2 + 2n(\mathbf{r}+\mathbf{r}')r' \right) dr'\\
  &= \frac1{R} \int 
  \delta(|\mathbf{r}'| - R)
  \left(\mathbf{r}'\cdot \nabla n(\mathbf{r}+\mathbf{r}') +
  2n(\mathbf{r}+\mathbf{r}') \right) \mathbf{dr}' \\
  &= \frac{2}{R}n_2(\mathbf{r}) + \frac{1}{R} \int 
  \delta(|\mathbf{r}'| - R)\mathbf{r}'\cdot \nabla n(\mathbf{r}+\mathbf{r}') \mathbf{dr}' \\
  \frac{dn_0(\mathbf{r})}{dR} &= \frac{1}{R} \int 
  \delta(|\mathbf{r}'| - R)\mathbf{r}'\cdot \nabla n(\mathbf{r}+\mathbf{r}') \mathbf{dr}' \\
  \frac{dn_1(\mathbf{r})}{dR} &=
    \frac1{4\pi R}\frac{dn_2(\mathbf{r})}{dR} - n_0(\mathbf{r}) \\
    &= n_0(\mathbf{r}) + \frac{1}{4\pi R^2} \int 
    \delta(|\mathbf{r}'| - R)\mathbf{r}'\cdot \nabla n(\mathbf{r}+\mathbf{r}') \mathbf{dr}' \\
  \frac{d\mathbf{n}_{V2}}{dR} &= \nabla n_2
\end{align}

The total force due to $\Phi_1$ is actually very easy to work out:
\begin{align}
  \frac{d A_{HS}^{(1)}}{d R} &=
  \int \mathbf{dr}' \left(\frac{n_0n_2}{1-n_3} - \frac{dn_0(\mathbf{r})}{dR}\ln(1-n_3) \right)
\end{align}
where you can see that $\frac{dn_0(\mathbf{r})}{dR}$ is zero for a
homogeneous system, and small for reasonably smooth density
distributions.  The next term from $\Phi_2$ is as follows:
\begin{align}
  \frac{d A_{HS}^{(2)}}{d R} &=
  \int \mathbf{dr}' \left(
    \frac{n_1 n_2 - \mathbf{n}_{V1} \cdot\mathbf{n}_{V2}}{(1-n_3)^2}
    n_2
    - \frac{n_0n_2 - \mathbf{n}_{V0} \cdot\mathbf{n}_{V2} - 2 n_1
      \frac{dn_2}{dR} + 2 \mathbf{n}_{V1} \cdot \nabla n_2 \cdot
    }{1-n_3}
  \right) \\
  &= \int \mathbf{dr}' \left(
    \frac{n_1 n_2 - \mathbf{n}_{V1} \cdot\mathbf{n}_{V2}}{(1-n_3)^2}
    n_2
    - \frac{n_0n_2 - \mathbf{n}_{V0} \cdot\mathbf{n}_{V2}
      - 2 n_1 \left(\frac2{R}n_2 \frac{dn_0}{dR}\right)
      + 2 \mathbf{n}_{V1} \cdot \nabla n_2 \cdot }{1-n_3}
  \right) \\
  &= \int \mathbf{dr}' \left(
    \frac{n_1 n_2 - \mathbf{n}_{V1} \cdot\mathbf{n}_{V2}}{(1-n_3)^2}
    n_2
    - \frac{n_0n_2 - \mathbf{n}_{V0} \cdot\mathbf{n}_{V2}
      - 4 n_0n_2 - 2 n_1\frac{dn_0}{dR}
      + 2 \mathbf{n}_{V1} \cdot \nabla n_2 \cdot }{1-n_3}
  \right) \\
  &= \int \mathbf{dr}' \left(
    \frac{n_1 n_2 - \mathbf{n}_{V1} \cdot\mathbf{n}_{V2}}{(1-n_3)^2}
    n_2
    + \frac{3n_0n_2 + \mathbf{n}_{V0} \cdot\mathbf{n}_{V2}
      + 2 n_1\frac{dn_0}{dR}
      - 2 \mathbf{n}_{V1} \cdot \nabla n_2 \cdot }{1-n_3}
  \right)
\end{align}

And the last term looks like this.  I should perhaps note that this
term should be replaced by a tensor version if want to examine
seriously localized density distributions.
\begin{align}
  A_{HS}^{(3)} &=
  \int \mathbf{dr}'\left(
     (n_2^3 - 3n_2\mathbf{n}_{V2} \cdot \mathbf{n}_{V2})
     \frac{
       n_3 + (1-n_3)^2 \ln(1-n_3)
     }{
       36\pi n_3^2\left( 1 - n_3 \right)^2
     }
  \right) \\
  \frac{d A_{HS}^{(3)}}{d R} &=
  \int \mathbf{dr}'
     \left[
       3(n_2^2 - \mathbf{n}_{V2} \cdot \mathbf{n}_{V2})
       \left(\frac2{R}n_2 + \frac{dn_0}{dR}\right)
       - 6n_2 \mathbf{n}_{V2} \cdot \nabla n_2
     \right]
     \frac{
       n_3 + (1-n_3)^2 \ln(1-n_3)
     }{
       36\pi n_3^2\left( 1 - n_3 \right)^2
     }
   \\
  &+
     (n_2^3 - 3n_2\mathbf{n}_{V2} \cdot \mathbf{n}_{V2}) n_2
     \frac{
       (1 - 2(1-n_3)\ln(1-n_3) - (1-n_3))n_3^2( 1 - n_3)^2
       -
       (n_3 + (1-n_3)^2 \ln(1-n_3))(-2 n_3^2( 1 - n_3) + 2n_3(1-n_3)^2)
     }{
       36\pi (n_3^2\left( 1 - n_3 \right)^2)^2
     }
     \\
     &=
     \int \mathbf{dr}'
     \left[
       (n_1n_2 - \mathbf{n}_{V1} \cdot \mathbf{n}_{V2})
       \left(2n_2 + \frac{1}{4\pi}\frac{dn_0}{dR}\right)
       - \frac{1}{2\pi}n_2 \mathbf{n}_{V2} \cdot \nabla n_2
     \right]
     \frac{
       n_3 + (1-n_3)^2 \ln(1-n_3)
     }{
       3 n_3^2\left( 1 - n_3 \right)^2
     }
   \\
  &+
     \frac{n_2(n_2^3 - 3n_2\mathbf{n}_{V2} \cdot \mathbf{n}_{V2})}{36\pi}
   \left( \frac{1}{(1-n_3)^2} + 2 \frac{\ln(1-n_3)}{n_3^2(1-n_3)} \right)
\end{align}

@@

\begin{align}
    \beta\frac{\delta A_{HS}}{\delta n_3(\mathbf{r}')} &=
    \frac{n_0(\mathbf{r}')}{1 - n_3(\mathbf{r}')}
    + \frac{n_1n_2 - \mathbf{n}_{V1}\cdot\mathbf{n}_{V2}}{(1 -
      n_3(\mathbf{r}'))^2}
    + \frac{n_2^3 -
      3n_2\mathbf{n}_{V2}\cdot\mathbf{n}_{V2}}{36\pi}
    %\left(
    %  \frac{1}{(1-n_3)^2} + 2\frac{\ln(1-n_3)}{n_2^2(1-n_3)}
    %\right)
    \\
    & \times \left(\frac{1}{n_3(1-n_3)^3} -\frac1{n_3^2(1-n_3)^2}  -
      \frac{1}{n_3^2(1-n_3)} - 2\frac{\ln(1-n_3)}{n_3^3}\right)
\end{align}
\end{widetext}

@@

\begin{widetext}
\section{The White Bear free energy for homogeneous systems}

Here is the White Bear free energy functional for the hard-sphere
fluid:~\cite{roth2002whitebear}
\begin{align}
  \frac{A_{HS}}{k_BT} &= \int \left\{
  -n_0 \ln\left( 1 - n_3\right)
  + \frac{n_1 n_2 - \mathbf{n}_{V1} \cdot\mathbf{n}_{V2}}{1-n_3}
  + (n_2^3 - 3 n_2 \mathbf{n}_{V2} \cdot \mathbf{n}_{V2}) \frac{
    n_3 + (1-n_3)^2 \ln(1-n_3)
  }{
    36\pi n_3^2(1-n_3)^2
  }
  \right\} \\
  &= \int \left\{
  -n \ln\left( 1 - \eta\right)
  + \frac{n \eta}{1-\eta}
  + n^3 (4\pi R^2)^3 \frac{
    \eta + (1-\eta)^2 \ln(1-\eta)
  }{
    36\pi \eta^2(1-\eta)^2
  }
  \right\} \\
  &= \int \left\{
  -n \ln\left( 1 - \eta\right)
  + \frac{n \eta}{1-\eta}
  + n^3 4\pi (4\pi R^3)^2 \frac{
    \eta + (1-\eta)^2 \ln(1-\eta)
  }{
    36\pi \eta^2(1-\eta)^2
  }
  \right\} \\
  &= \int \left\{
  -n \ln\left( 1 - \eta\right)
  + \frac{n \eta}{1-\eta}
  + n\eta^2 4\pi 9 \frac{
    \eta + (1-\eta)^2 \ln(1-\eta)
  }{
    36\pi \eta^2(1-\eta)^2
  }
  \right\} \\
  &= N \left\{
  -\ln\left( 1 - \eta\right)
  + \frac{\eta}{1-\eta}
  + \frac{
    \eta + (1-\eta)^2 \ln(1-\eta)
  }{
    (1-\eta)^2
  }
  \right\} \\
  &= N \left\{
  -\ln\left( 1 - \eta\right)
  + \frac{\eta}{1-\eta}
  + \frac{
    \eta
  }{
    (1-\eta)^2
  }
  + \ln(1-\eta)
  \right\} \\
  &= N \left\{
  \frac{\eta}{1-\eta}
  + \frac{
    \eta
  }{
    (1-\eta)^2
  }
  \right\} \\
  &= N
  \frac{\eta(1-\eta) + \eta
  }{
    (1-\eta)^2
  } \\
  &= N
  \frac{2\eta -\eta^2}{(1-\eta)^2}
\end{align}
This is odd, because Wikipedia says that for the Carnahan-Starling
equation of state the excess free energy is
\begin{align}
  A = Nk_BT\frac{4\eta -3\eta^2}{(1-\eta)^2}
\end{align}
\end{widetext}
